---
title: "HW9"
output: pdf_document
date: "2025-04-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup 1: Regression Tree (Car seat Data)
In the textbook, a classification tree is applied to the Carseats dataset from the ISLR2 package after converting Sales into a qualitative response variable. Now, we will seek to predict Sales using regression trees, treating the response as a quantitative variable.

## Questions
a. Split the data set into a training set and a test set.
```{r}
library(ISLR2)
summary(Carseats)

train_indices = sample(1:nrow(Carseats), nrow(Carseats) * 0.7)

train_data = Carseats[train_indices, ]
test_data <- Carseats[-train_indices, ]
```


b. Fit a regression tree to the training set. Plot the tree, and interpret the results. What test MSE do you obtain?
```{r}
library(tree)
tree_model <- tree(Sales ~ ., data = train_data)
plot(tree_model)
text(tree_model, pretty = 0)

predictions <- predict(tree_model, newdata = test_data)

mse <- mean((predictions - test_data$Sales)^2)
print(mse)
```


c. Use cross-validation in order to determine the optimal level of tree complexity. Does pruning the tree improve the test MSE?
```{r}
cv_result <- cv.tree(tree_model)
plot(cv_result$size, cv_result$dev, type = "b",
     xlab = "Tree Size", ylab = "Deviance (CV Error)")

best_size <- cv_result$size[which.min(cv_result$dev)]
cat("Optimal tree size:", best_size, "\n")

pruned_tree <- prune.tree(tree_model, best = best_size)
plot(pruned_tree)
text(pruned_tree, pretty = 0)

pred_pruned <- predict(pruned_tree, newdata = test_data)
mse_pruned <- mean((pred_pruned - test_data$Sales)^2)
cat("Test MSE (pruned tree):", mse_pruned, "\n")

mse_unpruned <- mean((predictions - test_data$Sales)^2)
cat("Test MSE (original tree):", mse_unpruned, "\n")

```


# Setup 2: Classification Tree (Orange Juice Data)
This problem involves the OJ data set which is part of the ISLR2 package.

## Questions
a. Create a training set containing a random sample of 800 observations, and a test set containing the remaining observations.
```{r}
library(ISLR2)
summary(OJ)

train_indices <- sample(1:nrow(OJ), 800)

train_data <- OJ[train_indices, ]
test_data <- OJ[-train_indices, ]
```

b. Fit a tree to the training data, with Purchase as the response and the other variables as predictors. Use the summary() function to produce summary statistics about the tree, and describe the results obtained. What is the training error rate? How many terminal nodes does the tree have?
```{r}
tree_oj <- tree(Purchase ~ ., data = train_data)

summary(tree_oj)
```
"LoyalCH", "PriceDiff", "DiscCH" were the most informative predictors for classifying Purchase.
Training error rate: 0.1638.
The tree ends in 8 terminal nodes.

c. Type in the name of the tree object in order to get a detailed text output. Pick one of the terminal nodes, and interpret the information displayed.
```{r}
tree_oj
```
13) LoyalCH > 0.75969 51 28.04 CH ( 0.92157 0.07843 ) *
Split condition: LoyalCH > 0.75969
n: 51 observations fall into this node
Deviance: 28.04
Predicted class: CH
Class probabilities:
CH: 92.16%
MM: 7.84%
Terminal node: * (no further splits)

d. Create a plot of the tree, and interpret the results.
```{r}
plot(tree_oj)
text(tree_oj, pretty = 0)
```
<!-- LoyalCH $\lt$ 0.5036 is most important split. Customers with LoyalCH $\lt$ 0.5036 are more likely to purchase MM, while those with LoyalCH $\geq$ 0.5036 lean toward CH. -->

e. Predict the response on the test data, and produce a confusion matrix comparing the test labels to the predicted test labels. What is the test error rate?
```{r}
pred_test = predict(tree_oj, newdata = test_data, type = "class")

conf_matrix = table(Predicted = pred_test, Actual = test_data$Purchase)
print(conf_matrix)

test_error = mean(pred_test != test_data$Purchase)
cat("Test Error Rate:", round(test_error, 4), "\n")
```

f. Apply the cv.tree() function to the training set in order to determine the optimal tree size.
```{r}
cv_oj <- cv.tree(tree_oj, FUN = prune.misclass)
print(cv_oj)
```

g. Produce a plot with tree size on the x-axis and cross-validated classification error rate on the y-axis.
```{r}
plot(cv_oj$size, cv_oj$dev, type = "b",
     xlab = "Tree Size (Terminal Nodes)",
     ylab = "CV Misclassification Error",
     main = "CV Error vs Tree Size")
```

h. Which tree size corresponds to the lowest cross-validated classification error rate?
```{r}
optimal_size <- cv_oj$size[which.min(cv_oj$dev)]
cat("Optimal tree size:", optimal_size, "\n")
```

i. Produce a pruned tree corresponding to the optimal tree size obtained using cross-validation. If cross-validation does not lead to the selection of a pruned tree, then create a pruned tree with five terminal nodes.
```{r}
pruned_tree <- prune.misclass(tree_oj, best = 8)
plot(pruned_tree)
text(pruned_tree, pretty = 0)
title("Pruned Classification Tree")
```

j. Compare the training error rates between the pruned and un-pruned trees. Which is higher?
```{r}
pred_train_full <- predict(tree_oj, newdata = train_data, type = "class")
train_error_full <- mean(pred_train_full != train_data$Purchase)
cat("Training Error (Unpruned Tree):", round(train_error_full, 4), "\n")

pred_train_pruned <- predict(pruned_tree, newdata = train_data, type = "class")
train_error_pruned <- mean(pred_train_pruned != train_data$Purchase)
cat("Training Error (Pruned Tree):", round(train_error_pruned, 4), "\n")
```

k. Compare the test error rates between the pruned and unpruned trees. Which is higher?
```{r}
Both the pruned and unpruned trees have the same training error rate.
```

