---
title: "HW9"
output: pdf_document
date: "2025-04-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup 1: Regression Tree (Car seat Data)
In the textbook, a classification tree is applied to the Carseats dataset from the ISLR2 package after converting Sales into a qualitative response variable. Now, we will seek to predict Sales using regression trees, treating the response as a quantitative variable.

## Questions
a. Split the data set into a training set and a test set.
```{r}
set.seed(1)
library(ISLR2)
n <- nrow(Carseats)
n_training <- round(n * 0.75)

index <- sample(n, size = n_training)
training_data <- Carseats[index, ]
test_data <- Carseats[-index, ]
```

b. Fit a regression tree to the training set. Plot the tree, and interpret the results. What test MSE do you obtain?
```{r}
library(tree)
tree_fit <- tree(Sales ~ ., data = training_data)
summary(tree_fit)
```
These six predictors are used to construct the tree: ShelveLoc, Price, Age, Advertising, Income, CompPrice.

The total number of terminal nodes is 16. The training MSE is 2.322.
```{r}
plot(tree_fit)
text(tree_fit, pretty = 0)

pred_test <- predict(tree_fit, newdata = test_data)
(test_MSE <- mean((test_data$Sales - pred_test)^2))
```


c. Use cross-validation in order to determine the optimal level of tree complexity. Does pruning the tree improve the test MSE?
```{r}
(cv_tree <- cv.tree(tree_fit, FUN = prune.tree))
plot(cv_tree$size, cv_tree$dev, xlab = "Tree Size", ylab = "Deviance", type = "b")
```
The size of optimum pruned tree:
```{r}
(opt_size <- cv_tree$size[which.min(cv_tree$dev)])
```
However, one can consider a tree of size 4, as it’s performance is almost the same as the optimum pruned tree.
```{r}
tree_pruned <- prune.tree(tree_fit, best = opt_size)
summary(tree_pruned)
```
Here is the test MSE:
```{r}
pred_test_pruned <- predict(tree_pruned, newdata = test_data)
(test_MSE <- mean((test_data$Sales - pred_test_pruned)^2))
```
The test MSE is increased after pruning. So, we do not see any advantage in pruning for this data set.

Let’s check the test error with the tree of size 4:
```{r}
tree_pruned <- prune.tree(tree_fit, best = 4)
pred_test_pruned <- predict(tree_pruned, newdata = test_data)
(test_MSE <- mean((test_data$Sales - pred_test_pruned)^2))
```
The test MSE is slightly lower than the best-pruned tree suggested by the cross-validation. However, its performance is still poor compared to the full tree.


# Setup 2: Classification Tree (Orange Juice Data)
This problem involves the OJ data set which is part of the ISLR2 package.

## Questions
a. Create a training set containing a random sample of 800 observations, and a test set containing the remaining observations.
```{r}
library(ISLR2)
library(tree)
summary(OJ)

set.seed(17)
train_indices <- sample(1:nrow(OJ), 800)

train_data <- OJ[train_indices, ]
test_data <- OJ[-train_indices, ]
```

b. Fit a tree to the training data, with Purchase as the response and the other variables as predictors. Use the summary() function to produce summary statistics about the tree, and describe the results obtained. What is the training error rate? How many terminal nodes does the tree have?
```{r}
tree_oj <- tree(Purchase ~ ., data = train_data)

summary(tree_oj)
```
"LoyalCH", "SalePriceMM", PriceDiff", "DiscCH" were the most informative predictors for classifying Purchase.
Training error rate: 0.185
The tree ends in 7 terminal nodes.

c. Type in the name of the tree object in order to get a detailed text output. Pick one of the terminal nodes, and interpret the information displayed.
```{r}
tree_oj
```
For node 10: $0.05 <$ LoyalCH $< 0.48$ and SalePriceMM $< 2.04$. 132 observations. 20% CH and 80% MM. Deviance = 132.So, the predicted value is MM.

d. Create a plot of the tree, and interpret the results.
```{r}
plot(tree_oj)
text(tree_oj, pretty = 0)
```
LoyalCH is most important split. Customers with LoyalCH $<$ 0.48 are more likely to purchase MM, while those with LoyalCH $\geq$ 0.48 lean toward CH. If $0.48 \geq$ LoyalCH $< 0.76$ and PriceDiff $\geq 0.08$ the predicted value is CH.

e. Predict the response on the test data, and produce a confusion matrix comparing the test labels to the predicted test labels. What is the test error rate?
```{r}
pred_test = predict(tree_oj, newdata = test_data, type = "class")

conf_matrix = table(Predicted = pred_test, Actual = test_data$Purchase)
print(conf_matrix)

test_error = mean(pred_test != test_data$Purchase)
cat("Test Error Rate:", round(test_error, 4), "\n")
```

f. Apply the cv.tree() function to the training set in order to determine the optimal tree size.
```{r}
cv_oj <- cv.tree(tree_oj, FUN = prune.tree)
print(cv_oj)
```

g. Produce a plot with tree size on the x-axis and cross-validated classification error rate on the y-axis.
```{r}
plot(cv_oj$size, cv_oj$dev, type = "b",
     xlab = "Tree Size (Terminal Nodes)",
     ylab = "Deviance",
     main = "CV Error vs Tree Size")
```

h. Which tree size corresponds to the lowest cross-validated classification error rate?
```{r}
optimal_size <- cv_oj$size[which.min(cv_oj$dev)]
cat("Optimal tree size:", optimal_size, "\n")
```

i. Produce a pruned tree corresponding to the optimal tree size obtained using cross-validation. If cross-validation does not lead to the selection of a pruned tree, then create a pruned tree with five terminal nodes.
```{r}
pruned_tree <- prune.tree(tree_oj, best = optimal_size)
plot(pruned_tree)
text(pruned_tree, pretty = 0)
title("Pruned Classification Tree")
summary(pruned_tree)
```

j. Compare the training error rates between the pruned and un-pruned trees. Which is higher?
Misclassification error rate: 0.185 for both.

k. Compare the test error rates between the pruned and unpruned trees. Which is higher?
```{r}
test_err = mean() 
```

Both the pruned and unpruned trees have the same test error rate.

