---
title: "Lab1"
author: "Atanu Giri"
date: "2025-01-21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1: Vector Manipulations
1. Create a numeric vector v of length 10 containing the first 10 positive integers.
2. Add these three numbers to vector v: 18, 20 and -15.
3. Remove the 3rd and 5th element of v.
4. Extract the 4th to 7th elements of v.
5. Replace the even numbers in v with NA.
6. Calculate the sum of all non-NA elements in v.
```{r}
(v = 1:10)
(v = c(v, 18, 20, -15))
(v = v[-c(3,5)])
(v[4:7])
v[v %% 2 == 0] = NA
v
(sum(v,na.rm=TRUE))
```

# Problem 2: Matrix Manipulations
1. Create a 3×3 matrix M with elements from 1 to 9 filled row-wise.
2. Extract the second row of the matrix.
3. Calculate the sum of each column in the matrix.
4. Add a new row [10, 11, 12] to the matrix.
5. Multiply the resulting matrix by 2.
6. Multiply the resulting matrix by a 3×4 matrix, where elements are randomly generated from the standard normal distribution, N(0,1). Call the new matrix P.
7. Find the determinant of P
```{r}
(M = matrix(1:9, nrow = 3, byrow = TRUE))
(M[2,])
(apply(M, 2, sum))
(M = rbind(M, c(10,11,12)))
(M = M*2)
M2 = matrix(rnorm(12), nrow = 3)
(P = M %*% M2)
det(P)
```

# Problem 3: Conditional Statements
1. Write an R script that checks whether a given number x is positive, negative, or zero and prints an appropriate message.
Example: If x = -3, the output should be: “x = -3 is negative.”
2. Create a function check_even_odd that takes a number as input and returns whether it is “Even” or “Odd”.
```{r}
x = -3

if (x > 0) {
  cat("x =", x, "is positive.\n")
} else if (x < 0) {
  cat("x =", x, "is negative.\n")
} else {
  cat("x =", x, "is zero.\n")
}
```

```{r}
check_even_odd = function(x)
  if (x %% 2 == 0){
    return("Even")  
  } else {
    return("Odd")
  }

result1 <- check_even_odd(4)
print(result1)
```

# Problem 4: Loops
## For Loop
1. Write a for loop to print the squares of numbers from 1 to 10.
2. Use a for loop to calculate the factorial of 5.
```{r}
for (i in 1:10){
  print(i^2)
}

s = 1
for (i in 1:5){
  s = s*i
}
print(s)
```
## While Loop
1. Write a while loop to print numbers from 1 to 10.
2. Write a while loop to calculate the sum of numbers from 1 to 50.
```{r}
i = 1
while (i < 11) {
  print(i)
  i = i+1
}

s = 0
i = 1
while (i < 51) {
  s = s+i
  i = i+1
}
print(s)
```

# Problem 5: User-Defined Functions
1. Write a function sum_and_product that takes two arguments and returns their sum and product as a list.
Example: For inputs 2 and 3, the output should be: list(sum = 5, product = 6).
2. Create a function generate_fibonacci that takes a single argument n and returns the first n Fibonacci numbers.
Example: For n = 5, the output should be: 1, 1, 2, 3, 5.
```{r}
sum_and_product = function(x,y){
  sum = x+y
  product = x*y
  return(list(sum = sum, product = product))
}

output = sum_and_product(2,3)
print(output)

generate_fibonacci <- function(n) {
  if (n <= 0) {
    stop("n must be a positive integer.")
  }
  
  # Initialize a vector to store Fibonacci numbers
  fibonacci <- numeric(n)
  
  if (n >= 1) fibonacci[1] <- 1
  if (n >= 2) fibonacci[2] <- 1
  
  for (i in 3:n) {
    fibonacci[i] <- fibonacci[i - 1] + fibonacci[i - 2]
  }
  
  return(fibonacci)
}

output = generate_fibonacci(5)
print(output)
```