---
title: "Lab2: Data Handling"
output: pdf_document
date: "2025-02-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This exercise involves the Boston housing data set. Write R codes to answer the following questions.

1. To begin, load in the Boston data set. The Boston data set is part of the MASS library in R. Per capita crime rate is the response variable. Print the first few observations of the dataset

   How many rows are in this data set? How many columns? What do the rows and columns represent (check R help)?
```{r}
library(MASS)
head(Boston)

dim(Boston)
```

-There are 506 rows and 14 columns.\
-Each row corresponds to a specific neighborhood or district within the Boston area.\
-Each column represents a different attribute related to housing, crime rates, and economic factors.\

2. Make some pairwise scatterplots of the predictors (columns) in this data set. Describe your findings.
```{r}
library(ggplot2)
library(GGally)
ggpairs(Boston)
```
As percentage of lower-income households (lstat) increases, median home values (medv) tend to decrease.
More rooms in a house are associated with a lower percentage of low-income residents and higher median home values.

Higher tax rates are associated with higher pupil-teacher ratios.
Higher industrialization correlates with higher nitrogen oxides concentration.

Some variables, such as medv and lstat, or nox vs distances to employment centres, show a curved pattern rather than a straight-line trend, suggesting a non-linear relationship.

3. Are any of the predictors associated with the per capita crime rate? If so, explain the relationship.
```{r}
sort(apply(Boston, 2, function(x) cor(Boston[, 1], x)))
```

4. Find the summary of each predictor. Check the top five observations with the highest per capita crime rate.
```{r}
summary(Boston)
sorted_data = sort(Boston[,1], decreasing = TRUE, index.return = TRUE)
data_5 = sorted_data$ix[1:5]
data_5_Boston = Boston[data_5,]
```

5. How many of the census tracts (observations) in this data set bound the Charles river?
```{r}
sum(Boston$chas == 1)
```

6. What is the median pupil-teacher ratio among the towns in this data set?
```{r}
median(Boston$ptratio)
```

7. Which census tract of Boston has the lowest median value of owner-occupied homes? Print the corresponding row.
```{r}
Boston[which.min(Boston$medv),]
```

8. In this dataset, how many of the census tracts average more than eight rooms per dwelling? Find the summary of the census tracts that average more than eight rooms per dwelling.
```{r}
sum(Boston$rm > 8)
summary(Boston[Boston$rm > 8,])
```

9. Save this dataset in an Excel or CSV file.
```{r}
write.csv(Boston, file = "Boston.csv", row.names = FALSE)

library(openxlsx)
write.xlsx(x = Boston, file = 'Boston_data.xlsx')
```
